# Reproducible Research: Peer Assessment 1


## Load libraries and Set options
```{r}
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

## Loading and preprocessing the data

1. Load the data (i.e. read.csv())

```{r}
data.act <- read.csv("activity.csv", header = TRUE, sep =",", quote = "\"", na.strings = "NA")
```


2. Process/transform the data (if necessary) into a format suitable for your analysis
```{r}
dates <- data.act$date
dates <- as.Date(as.character(dates), "%Y-%m-%d")
str(data.act)
#plot(dates, data.act$steps)
```


## What is mean total number of steps taken per day?

To get the mean total number of steps taken per day, we get the total number of steps taken per day
```{r}
steps.sum <- with(data.act, tapply(steps, dates, sum, na.rm = T))
str(steps.sum)
summary(steps.sum)
```


1. Make a histogram of the total number of steps taken each day

Here is a histogram of the total number of steps taken per day
```{r}
steps.perday <- data.frame(date = names(steps.sum), steps = steps.sum)
hist(steps.perday$steps, breaks = 20, main = "total number of steps taken each day", xlab = "number of steps", ylab = "frequency")
```


2. Calculate and report the mean and median total number of steps taken per day

--> The mean of total number of steps taken per day is:
```{r}
mean(steps.perday$steps)
```

--> The median of total number of steps taken per day:
```{r}
median(steps.perday$steps)
```

## What is the average daily activity pattern?

To get the average daily activity pattern, let's examine the mean steps per 5-min interval in a day across all all days --
```{r}
interval.mean <- with(data.act, tapply(steps, interval, mean, na.rm = T))
head(interval.mean)
steps.perint <- data.frame(interval = as.numeric(names(interval.mean)), steps = interval.mean)
```

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

Here's a time series plot of that average per 5-min interval - 
```{r}
qplot(interval, steps, data = steps.perint, geom = "line", main = "average daily activity pattern", xlab = "5-min interval", ylab = "average steps, across all days")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

--> The 5-min interval which contains, on the average, the maximum number of steps:
```{r}
max <- max(steps.perint$steps)
steps.perint[which(steps.perint$steps == max),]
```

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

--> The number of missing values for the steps in the dataset is:
```{r}
sum(is.na(data.act$steps))
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

--> I propose that we use the mean across all days for that interval as the value to substitute the missing values in the dataset. 

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

--> Here's the code that substitutes the mean steps across all days for missing steps per interval:
```{r}
act.cleaned <- data.act
# if steps is NA, use mean of steps for that interval
for (i in 1:nrow(data.act)) {
     if (is.na(data.act$steps[i])) {
          act.cleaned$steps[i] <- interval.mean[as.character(data.act$interval[i])]
     }
}
head(act.cleaned)
tail(act.cleaned)
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

With the missing values substituted as proposed, here's the histrogram again of the total number of steps taken each day
```{r}
steps.sumcleaned <- with(act.cleaned, tapply(steps, date, sum, na.rm = T))
steps.perdaycleaned <- data.frame(date = names(steps.sumcleaned), steps = steps.sumcleaned)

par(mfrow = c(2, 1))
hist(steps.perday$steps, breaks = 20, main = "total number of steps taken each day, with missing values", xlab = "number of steps", ylab = "frequency")
hist(steps.perdaycleaned$steps, breaks = 20, main = "total number of steps taken each day, cleaned data", xlab = "number of steps", ylab = "frequency")
```


--> The means and medians of the dataset with missing values compared to the cleaned dataset are:
```{r}
c(mean(steps.perday$steps),median(steps.perday$steps))
c(mean(steps.perdaycleaned$steps),median(steps.perdaycleaned$steps))
```

Do these values differ from the estimates from the first part of the assignment? 

--> Yes

What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
summary(steps.perday$steps)
summary(steps.perdaycleaned$steps)
```

--> The summary statistics increased.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

--> Let's create daytype as the factor that will contain whether the data was on a "weekday" or a "weekend"
```{r}
dayofweek <- weekdays(dates)
daytype = c()
for (i in 1:length(dayofweek)) { 
     daytype[i] <- if (dayofweek[i] == "Saturday" | dayofweek[i] == "Sunday") "weekend" else "weekday" 
     }
head(daytype)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data. Your plot will look different from the one above because you will be using the activity monitor data. Note that the above plot was made using the lattice system but you can make the same version of the plot using any plotting system you choose.

Here's the panel plot -
```{r}
data.weekday <- subset (act.cleaned, daytype == "weekday")
data.weekend <- subset (act.cleaned, daytype == "weekend")

interval.mean.day <- with(data.weekday, tapply(steps, interval, mean))
steps.perint.day <- data.frame(interval = as.numeric(names(interval.mean.day)), steps = interval.mean.day)

interval.mean.end <- with(data.weekend, tapply(steps, interval, mean))
steps.perint.end <- data.frame(interval = as.numeric(names(interval.mean.end)), steps = interval.mean.end)

require(gridExtra)
plotday <- qplot(interval, steps, data = steps.perint.day, geom = "line", main = "weekdays", xlab = "5-min interval", ylab = "ave num of steps")
plotend <- qplot(interval, steps, data = steps.perint.end, geom = "line", main = "weekends", xlab = "5-min interval", ylab = "ave num of steps")
grid.arrange(plotday, plotend, nrow=2)
```
